---
title: "Machine Learning Project Assignment"
author: "Amit Kumar"
date: "2 August 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load required packages

library(caret)
library(rattle)

### Load training and test data 

TrainData <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"),header=TRUE)
dim(TrainData)

TestData <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"),header=TRUE)
dim(TestData)

head(TestData)

str(TrainData)

```


#Observations
1.The training data set is made of 19622 observations on 160 columns.
2.And,We can notice that many columns have NA values or blank values on almost every observation. So we will remove them, because they will not produce any information. 
3.The first seven columns give information about the people who did the test, and also timestamps. We will not take them in our model.

```{r}

# Here we get the indexes of the columns having at least 90% of NA or blank values on the training dataset
indColToRemove <- which(colSums(is.na(TrainData) |TrainData=="")>0.9*dim(TrainData)[1])
# remove these blank variables from both training datasets
TrainDataClean <- TrainData[,-indColToRemove]
##### removing first 7 columns
TrainDataClean <- TrainDataClean[,-c(1:7)]
dim(TrainDataClean)


# We do the same operation on test dataset
indColToRemove <- which(colSums(is.na(TestData) |TestData=="")>0.9*dim(TestData)[1]) 
TestDataClean <- TestData[,-indColToRemove]
TestDataClean <- TestDataClean[,-1]
dim(TestDataClean)

#Note: After cleaning, the new training data set has only 53 columns

# Here we create a partition of the traning data set 
set.seed(12345)
inTrain1 <- createDataPartition(TrainDataClean$classe, p=0.75, list=FALSE)
Train1 <- TrainDataClean[inTrain1,]
Test1 <- TrainDataClean[-inTrain1,]
dim(Train1)
dim(Test1)

```

# Next Steps

Now, let us try following 3 different models:
1. Classification Trees
2. Random Forest
3. Gradient Boosting Method

In order to limit the effects of overfitting, and improve the efficicency of the models, we will usecross-validation technique.
We will use 5 folds (usually, 5 or 10 can be used, but 10 folds gives higher run times with no significant increase of the accuracy).

```{r}

trControl <- trainControl(method="cv", number=5)
model_CT <- train(classe~., data=Train1, method="rpart", trControl=trControl)

print(model_CT)
fancyRpartPlot(model_CT$finalModel)

trainpred <- predict(model_CT,newdata=Test1)

confMatCT <- confusionMatrix(Test1$classe,trainpred)

# display confusion matrix and model accuracy
confMatCT$table

confMatCT$overall[1]


```


# Next, Let us try to fit Random Forest Model

```{r}

model_RF <- train(classe~., data=Train1, method="rf", trControl=trControl, verbose=FALSE)
print(model_RF)


plot(model_RF,main="Accuracy of Random forest model by number of predictors")

trainpred <- predict(model_RF,newdata=Test1)

confMatRF <- confusionMatrix(Test1$classe,trainpred)

# display confusion matrix and model accuracy
confMatRF$table

confMatRF$overall[1]

model_RF$finalModel$classes

plot(model_RF$finalModel,main="Model error of Random forest model by number of trees")

```

#Observation from Random Forest

With random forest, we reach an accuracy of 99.3% using cross-validation with 5 steps. This is very good. But let's see what we can expect with Gradient boosting.

We can also notice that the optimal number of predictors, i.e. the number of predictors giving the highest accuracy, is 27. There is no significal increase of the accuracy with 2 predictors and 27, but the slope decreases more with more than 27 predictors (even if the accuracy is still very good). The fact that not all the accuracy is worse with all the available predictors lets us suggest that there may be some dependencies between them.

At last, using more than about 30 trees does not reduce the error significantly.


#Train with gradient boosting method

```{r}

model_GBM <- train(classe~., data=Train1, method="gbm", trControl=trControl, verbose=FALSE)
print(model_GBM)
plot(model_GBM)

trainpred <- predict(model_GBM,newdata=Test1)

confMatGBM <- confusionMatrix(Test1$classe,trainpred)
confMatGBM$table

confMatGBM$overall[1]

```

#Conclusion
This shows that the random forest model is the best one. 
We will then use it to predict the values of classe for the test data set.


```{r}

FinalTestPred <- predict(model_RF,newdata=TestDataClean)
FinalTestPred

```
## R Markdown
